---
title: "Fire Explorer Report"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data Cleaning and Processsing 
### Install Pakcage
We use function package.require to install and load packages. If the packages do not exist, it will be installed automatically.
```{r}
#### package.require #####
#### @param: packages install and load package ####
package.require <- function(packages){
  package.local <- as.data.frame(installed.packages()[,c("Package","Version")])
  package.need.install <- setdiff(packages,
                                  as.character(package.local$Package))
  if(length(package.need.install) != 0){
    install.packages(package.need.install)
  }
  for(package in packages){
    require(package,character.only = TRUE)
  }
}

package.require(c("dplyr", "foreign", "readr", "data.table", "lubridate"))
```


### Read Data
We build the file.read function to read two extension file dbf, and txt.
- Read the codelookup
```{r}

#### get all the file directory in fire ####
path <- paste0(getwd(),"/",unlist(lapply(list.dirs("Fire"), function(x) paste0(x,"/",list.files(x)))))

#######                     fire.read                   #########
####### @param: path.dir the file path list or vector   #########
####### return: the list of data.table                  #########
fire.read <- function(path.dir){
  df.list <- list()
  file.n <- length(path.dir)
  year <-  sapply(path.dir, function(x) unlist(strsplit(x,"/"))[4])
  for(i in 1:file.n){
    cat(paste0("File in ", path.dir[i]," is reading"),"\n")
    # if the file extension is dbf, use the read.dbf to read file
    if(grepl("dbf", tolower(path.dir[i])))
      ds <- data.table(read.dbf(path.dir[i], as.is=TRUE))
    else
      # else use the readr to read file
      ds <- read_delim(path.dir[i], delim = "^", col_names = TRUE)
    colnames(ds) <- tolower(colnames(ds))
    df.list[[i]] <- ds
  }
  # name list using year
  names(df.list) <- year
  gc()
  return(df.list)
}

#### read all the codelookup, and add variable year  ####
codelookup <- fire.read(path[grepl("codelookup", path)])
codelookup <- lapply(2006:2015, function(x)
  codelookup[[x-2005]] %>% mutate(year = x)) %>%
  rbindlist

#### get the description of inc_type in the codelookup #### 
inc_type.convert.df <- codelookup %>%
  filter(fieldid == "INC_TYPE") %>%
  mutate(inc_type = as.numeric(code_value)) %>%
  select(-fieldid, -code_value)

#### get the abbreviation of the state
state_code <- codelookup %>%
  filter(fieldid == "STATE" & !is.na(code_value)) %>%
  mutate(state = code_value)  %>%
  select(state, code_descr) %>%
  unique
```


- Read the basicincident
When we read the basicincident, we need do some transform, format all the key variable in suitable format and derive the day, weekday, month, year from the vairbale INC_DATE. There are two format in INC_DATE, at first, we transform it into character, if the length of INC_DATE < 7, we should paste 0 0 in front of the INC_DATE, then we use funciton lubridate::mdy function transorm it into date format.
```{r}

####                 key.transoform                      #####
#### THe function to make the primary key unified format ####
key.transform <- function(df){
  df %>%
    mutate(fdid = as.numeric(fdid)) %>%
    mutate(inc_date = as.numeric(inc_date)) %>%
    mutate(inc_no = as.numeric(inc_no)) %>%
    mutate(exp_no = as.numeric(exp_no))
}
####                 inc_date.transfrom                    ####
#### To derive the year, month, day, weekday from inc_date ####
inc_date.transfrom <- function(df){
  df %>%
    mutate(inc_date = mdy(ifelse(nchar(inc_date) == 7, paste0(0,inc_date),as.character(inc_date)))) %>%
    mutate(year = as.integer(year(inc_date))) %>%
    mutate(month = month(inc_date,label = TRUE, abbr = TRUE)) %>%
    mutate(day = day(inc_date)) %>%
    mutate(weekday = wday(inc_date, label = TRUE, abbr = FALSE))
}
####              basicincident.stat                   #####
####       calculate summarized statistics             #####
basicincident.stat <- function(df) {
  df %>%
    mutate(hh = as.integer(as.numeric(alarm)%%10^4/100)) %>%
    group_by(state, year, month, weekday, day, hh, inc_type) %>%
    summarise(n = n(),
              ff_death = sum(ff_death, na.rm = TRUE),
              oth_death = sum(oth_death, na.rm = TRUE),
              ff_inj = sum(ff_inj, na.rm = TRUE),
              oth_inj = sum(oth_inj, na.rm = TRUE))
}


basicinciden.list <- fire.read(path[grepl("basicincident", path)]) %>%
  lapply(function(df) df %>% 
           select(state, fdid, inc_date, inc_no, exp_no, inc_type, alarm, ff_death, oth_death, ff_inj, oth_inj) %>% key.transform %>% inc_date.transfrom
           )

Fire.Stat <- basicinciden.list %>%
  lapply(function(df) df %>% basicincident.stat) %>%
  rbindlist %>%
  mutate(inc_type = as.numeric(inc_type)) %>%
  inner_join(inc_type.convert.df, by = c("inc_type", "year"))
```

- Read teh hazmat information
```{r}
#### read teh hazmat information ####
hazmat <- fire.read(path[grepl("hazmat", path) & !grepl("hazmatequipinvolved", path)]) %>%
  lapply(function(df) df %>% key.transform %>% inc_date.transfrom)

  
for(i in 1:10){
  hazmat[[i]] <- hazmat[[i]] %>%
    inner_join(basicinciden.list[[i]] %>% 
                 select(state, fdid, inc_date, inc_no, exp_no, inc_type) %>%
                 mutate(inc_type = as.numeric(inc_type))) %>% 
    inner_join(inc_type.convert.df, by = c("inc_type", "year"))
}
```

- load the fifty state map
```{r}

load("fifty_states.RData")
  
state.position <- fifty_states %>%
  group_by(id) %>%
  summarise(latitude = mean(lat), longitude  = median(long))
```


## Statistic Analysis Fire

### Calcuate the Number of Fire in state by year.
```{r}

state.df <- Fire.Stat %>%
  group_by(state, year) %>%
  summarise(sum = sum(n)) %>%
  inner_join(state_code) %>%
  ungroup %>%
  mutate(id = tolower(code_descr), 
         state = code_descr) %>%
  inner_join(state.position)

save(state.df, file = "state.df")
```

### Calculate the number of different fires with year
```{r}

####       Calculate the number of occurrences of different fires per year     #####
Fire.year.stat <- Fire.Stat %>%
  group_by(year, inc_type, code_descr) %>%
  summarise(n = n())
#### save the result into .RData  ####
save(Fire.year.stat, file = "Fire.year.stat.RData")

####    Calculate the number of occurrences of different fires per hour     #####
Fire.time.stat <- Fire.Stat %>%
  group_by(year, hh, inc_type, code_descr) %>%
  summarise(n = n())
#### save the result into .RData  ####
save(Fire.time.stat, file = "Fire.time.stat.RData")
```


###    Calculate the number of occurrences of person death in different types of fires per hour
```{r}
death.month.week.stat <- hazmat %>%
  rbindlist %>%
  group_by(month, weekday, inc_type, code_descr) %>%
  summarise(death.num = sum(haz_death, na.rm = TRUE))

save(death.month.week.stat, file = "death.month.week.stat.RData")

death.fire <- death.month.week.stat %>%
  ungroup %>%
  filter(death.num > 0) %>%
  select(inc_type,code_descr) %>%
  distinct
save(death.fire, file = "death.fire.RData")
```
